Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-05-19 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Major modifications to account for new model selection

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "RibozeroBatch", "LibraryBatch", "FlowcellBatch", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('MSSM_RNA_ACC_BP_24',

                'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 
                'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220', 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133',

                'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 
                'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56')

# Outliers from Menahce's DLPFC analysis syn3493755
# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
#   'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218', 
#   'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
#   'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
#   'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
#   'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
#   'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
#   'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
#   'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537',  'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878',
#   'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_361',
#   'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047', 'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247',
#   'PITT_RNA_BP_PFC_1282', 'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')

@@

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r temp, include = F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. 
1. "Dx.Tissue" is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue)
2. Adding Instituion and Gender a priori to variable selection as they are correlated with Dx.Tissue
```{r iterative.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender');

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
residCovars = setdiff(residualSigCovars$significantCovars, c(postAdjustCovars, 'Individual_ID'))
covariatesEffects = residualSigCovars$Effects.significantCovars[residCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

loopCount = 0 
while(length(residualSigCovars$significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:',
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects and Individual_ID is chosen as random effect'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
  correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

  # Re-calculate voom weights with correlation of random effects
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                              design=DM1$design, plot=F,
                              block = COVARIATES$Individual_ID, 
                              correlation = correlation$cor)

  # Fit linear model using new weights and new design
  VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
  ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, 'Individual_ID'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F], 
                                                'adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  covariatesEffects = residualSigCovars$Effects.significantCovars[setdiff(residualSigCovars$significantCovars, 
                                                                          c(postAdjustCovars, 'Individual_ID'))]

  postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}

modelStr <- paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=', '),
                  'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES, 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

>>>>

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "PercentAligned", "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
# Based on PC
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

# Based on number of gene outliers
indToRemove = c(indToRemove, "MSSM_RNA_ACC_BP_15", "MSSM_RNA_ACC_175", "MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", 
                "MSSM_RNA_ACC_285", "MSSM_RNA_ACC_344", "MSSM_RNA_ACC_376", "MSSM_RNA_ACC_374", "PENN_RNA_ACC_003", 
                "PITT_RNA_ACC_539", "MSSM_RNA_PFC_201", "MSSM_RNA_PFC_234", "MSSM_RNA_PFC_355", "MSSM_RNA_PFC_389",
                "MSSM_RNA_PFC_80", "MSSM_RNA_PFC_83", "PENN_RNA_PFC_22", "PENN_RNA_PFC_59", "PENN_RNA_PFC_62", 
                "PENN_RNA_PFC_80", "PENN_RNA_PFC_105", "PITT_RNA_PFC_1472", "PITT_RNA_PFC_566", "PITT_RNA_PFC_727", 
                "PITT_RNA_PFC_802")

# # Outliers from Menahcem's DLPFC analysis syn3493755
# outliers = c('MSSM_RNA_BP_PFC_1', 'MSSM_RNA_PFC_21', 'MSSM_RNA_BP_PFC_3', 'MSSM_RNA_PFC_103', 
#              'MSSM_RNA_PFC_105', 'MSSM_RNA_PFC_115', 'MSSM_RNA_PFC_169',  'MSSM_RNA_PFC_17',
#              'MSSM_RNA_PFC_18', 'MSSM_RNA_PFC_19', 'MSSM_RNA_PFC_213', 'MSSM_RNA_PFC_214', 
#              'MSSM_RNA_PFC_215', 'MSSM_RNA_PFC_216', 'MSSM_RNA_PFC_217', 'MSSM_RNA_PFC_218',
#              'MSSM_RNA_PFC_219', 'MSSM_RNA_PFC_222', 'MSSM_RNA_PFC_24', 'MSSM_RNA_PFC_252', 
#              'MSSM_RNA_PFC_259', 'MSSM_RNA_PFC_263', 'MSSM_RNA_PFC_267', 'MSSM_RNA_PFC_275',
#              'MSSM_RNA_PFC_307', 'MSSM_RNA_PFC_322', 'MSSM_RNA_PFC_325', 'MSSM_RNA_PFC_27', 
#              'MSSM_RNA_PFC_332', 'MSSM_RNA_PFC_333', 'MSSM_RNA_PFC_334', 'MSSM_RNA_PFC_335',
#              'MSSM_RNA_PFC_340', 'MSSM_RNA_PFC_35', 'MSSM_RNA_PFC_362', 'MSSM_RNA_PFC_372', 
#              'MSSM_RNA_PFC_373', 'MSSM_RNA_PFC_385', 'MSSM_RNA_PFC_393', 'MSSM_RNA_PFC_395',
#              'MSSM_RNA_PFC_49', 'MSSM_RNA_PFC_54', 'MSSM_RNA_PFC_70', 'MSSM_RNA_PFC_76', 
#              'MSSM_RNA_PFC_77', 'MSSM_RNA_PFC_78', 'MSSM_RNA_PFC_88', 'MSSM_RNA_PFC_93',
#              'FAKE_DLPFC_CMC_MSSM_426', 'PENN_RNA_PFC_29', 'FAKE_DLPFC_CMC_PENN_055', 
#              'PENN_RNA_PFC_75', 'PENN_RNA_PFC_110', 'FAKE_DLPFC_CMC_PENN_106', 'FAKE_DLPFC_CMC_PENN_107',
#              'FAKE_DLPFC_CMC_PENN_108', 'MSSM_RNA_PFC_14', 'MSSM_RNA_BP_PFC_21', 'MSSM_RNA_PFC_293', 
#              'MSSM_RNA_PFC_349', 'MSSM_RNA_PFC_353', 'MSSM_RNA_PFC_363', 'MSSM_RNA_PFC_369',
#              'PENN_RNA_PFC_82', 'PENN_RNA_PFC_89', 'PITT_RNA_PFC_1159', 'PITT_RNA_PFC_1296', 'PITT_RNA_PFC_537', 
#              'PITT_RNA_PFC_567', 'PITT_RNA_PFC_871', 'PITT_RNA_PFC_878', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139',
#              'MSSM_RNA_PFC_210',  'MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_278', 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_348',
#              'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_56', 'PITT_RNA_BP_PFC_10003', 'PITT_RNA_BP_PFC_1047',
#              'PITT_RNA_BP_PFC_1086', 'PITT_RNA_BP_PFC_1092', 'PITT_RNA_BP_PFC_1247', 'PITT_RNA_BP_PFC_1282', 
#              'PITT_RNA_BP_PFC_1324', 'PITT_RNA_BP_PFC_1374', 'PITT_RNA_BP_PFC_1391', 'PITT_RNA_BP_PFC_686')

@@

Based on the expression pattern following `r length(indToRemove)` samples were tagged as outliers based on their mrna expression: `r paste(indToRemove, collapse = ', ')`

@@

### Identify gene outliers
Set outlier expression of genes in each samples to NA
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = NA

# Distribution of fraction NA genes in samples
tmp.frac = apply(log.mat, 2, sum)/dim(LOG.CPM)[1]
writeLines('Distribution of fraction NAs in samples')
quantile(tmp.frac, seq(0, 1, 0.1)) %>%
  round(3) 
top.nsamples = names(tmp.frac)[tmp.frac >= 0.02]
```
Only `r length(top.nsamples)` samples have more than 2% of NA genes. They are: `r paste(top.nsamples, collapse = ', ')`

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Function to convert NA values to zero
changeNAtoZero <- function(mtrx){
  mtrx[is.na(mtrx)] = 0
  return(mtrx)
}

# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
CQN.GENE_EXPRESSION$E[is.na(PROCESSED_COUNTS$filteredExprMatrix$counts)] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
plotdata$label = plotdata$SampleID
plotdata$label[!(plotdata$label %in% top.nasamples)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E %>% changeNAtoZero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(Dx+.~Tissue, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E %>% changeNAtoZero, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

### Normalisation
Based on different model selection methodology (as seen here syn8468116, syn9772932, and syn9774491), following variables are used as covariates: Institution, Gender, RIN, and IntronicRate

NOTE: Dx.Tissue is chosen as the primary variable of interest (i.e., covariate selection is conditioned on diagnosis and tissue), and Individual\_ID is chosen as the random effect

```{r custom.normalisation, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'IntronicRate')

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DESIGN = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero, 
                            design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts %>% changeNAtoZero,   
                            design = DESIGN, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DESIGN[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION %>% changeNAtoZero)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

p = DE %>%
  dplyr::filter(ref.Dx != 'Other', to.Dx != 'Other', ref.Dx != 'ALL', to.Dx != 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Dx+.~to.Dx+to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

p = DE %>%
  dplyr::filter(ref.Dx == 'ALL', to.Dx == 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Tissue+.~to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between DLPFC and ACC')
p
```
